---
title: "Homework 1 (Deadline: 11/02/2024)"
author: "Your Name Here"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

## Setup

You are free to use any function/package you want (from base R or any package) to complete this assignment. Additionally, people working in Python are free to submit their work in a juptyer notebook. 

For people doing it in R, please write your code in the appropriate code chunks below. You can submit a *knitted* file (html or pdf), or an `.Rmd` file. For more information on how to knit a file, please refer to the [R Markdown documentation](https://rmarkdown.rstudio.com/lesson-1.html).

```{r}
require(tidyverse)

# Load the data
federalist <- read_csv("data/federalist.csv")
```

## Task 1 Repeat the Authorship Analysis for All Disputed Federalist Papers

<ol>

<li>Identify Disputed Papers

Look at the `author` column to find rows labeled "HAMILTON OR MADISON"

<li> Author Frequencies

Pick a set of words (e.g., the three words that we used: “man,” “by,” “upon”) + 5 other words that you think are good predictors of authorship.

For each author (Hamilton, Madison, and Jay), compute their **word probabilities** (the ralitive rates) $\hat{\mu}_h, \hat{\mu}_m, \hat{\mu}_j$ (with Laplace+1 smoothing).

<li> For Each Disputed Document

Compute the multinomial likelihood of that document under each author’s distribution, for example, using `dmultinom(...)` function.

Which author has the largest likelihood?

<li> Either plot the results or create a table that shows the likelihood of each author for each disputed document.

</ol>

```{r}

# Your code here

```

## Task 2 Logged Odds Differences for Bigrams

<ol>

<li> Tokenize text into bigrams.

<li> Either remove every bigram that contains a stopword, or remove stopwords before tokenizing into bigrams.

<li> Remove the bigrams that occur only once.

<li> Separate the bigrams by author (focus only Hamilton and Madison).

<li> Compute the relative frequencies of bigrams for each author.

<li> Compute the log odds for each bigram.

\[
\log O_b^i
\;=\;
\log\!\biggl(\frac{f_b^i}{1 - f_b^i}\biggr),
\]

where $f_b^i$ is the relative frequency of bigram $b$ in author $i$.

<li> Compute the log odds differences for each bigram.

$$
\Delta_{bigram} = \log(\frac{f_b^i}{1 - f_b^i}) - \log(\frac{f_b^j}{1 - f_b^j})
$$

where $i$ and $j$ are the authors being compared.

<li> Plot the log odds differences of top discriminative bigrams.

</ol>

```{r}

# Your code here

```

